The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-13 paper/iphas-dr2.tex
Processed feedback from YU

Although the IPHAS data do contain a significant number of 
such inter-field repeat measurements,
we have found the information contained
in these regions to be insufficient
to constrain the calibration parameters.
This is because photometry at the extreme edges of the WFC
-- where neighbouring fields overlap -- 
is prone to systematics at the level of 1--2\%.
The cause of these errors is thought to include 
the use of twilight sky flats in the pipeline,
which are known to be imperfect for calibrating stellar photometry 
due to stray light and vignetting \citep[e.g.][]{Manfroid1995}.
Moreover, the illumination correction in the overlap regions
is more affected by a radial geometric distortion in the WFC,
which causes the pixel scale to increase as the edges 
are approached \citep{Gonzalez-Solares2011}.
Although these systematics are reasonably small within a single field,
they can combine, during the global calibration routine, to cause artificial zeropoint gradients 
across the survey unless controlled by
%when they are used to constrain a global calibration
%without 
other external constraints.

For these reasons, we have decided not to depend
on an ubercalibration-type scheme alone,
and have opted to involve an external reference survey
-- where available --
to bring the majority of our data onto a uniform calibration.
%This is explained next.

@@

We have been able to benefit from the
AAVSO Photometric All-Sky Survey
(APASS; http://www.aavso.org/apass)
to bring 
%the vast majority - exaggerating a bit? 
most of the survey 
onto a uniform scale.
Since 2009,
APASS has been using two 20~cm-astrographs
to survey the entire sky down to $\sim$17th magnitude
in five filters which include Sloan $r$ and $i$ \citep{Henden2012}.
The most recent catalogue available 
at the time of preparing this work was APASS DR7,
which provides a good coverage across $\sim$half
of the IPHAS footprint.
The overlap regions are shown in Fig.~\ref{fig:apass_r}.
The photometric accuracy of APASS is currently estimated 
to be at the level of 3\% (Henden, private communication),
which is significantly better 
than the original nightly calibrations of IPHAS,
for which we estimate the $1\sigma$-uncertainty to be $\sim$10\%.
APASS achieves its uniform accuracy 
by measuring each star at least two times in photometric conditions
-- along with ample standard fields --
using the large $3\times3$ square degrees field of view of its detectors.

With the aim of bringing IPHAS to a similar accuracy of $\sim$3\%,
we used the APASS catalogue to identify and adjust the calibration of all IPHAS fields 
which showed a magnitude offset larger than 3\% against APASS.  Experience of re-running
the calibration and testing the results showed us that it was inadvisable to more
finely tune the match for IPHAS data obtained in what were generally the most photometric
nights.  To this end,
the $r$- and $i$-band detection tables of each IPHAS field
were cross-matched against the APASS DR7 catalogue 
using a maximum matching distance of 1~arcsec.
The magnitude range was limited to
$13<r_{\rm APASS}<16.5$ and $12.5<i_{\rm APASS}<16.0$
in order to avoid sources 
brighter than the IPHAS saturation limit on one hand, 
and to avoid sources near the faint detection limit of APASS 
on the other.

The resulting set of $\sim$220\,000 cross-matched stars were then used 
to derive APASS-to-IPHAS magnitude transformations
using a linear least-squares fitting routine, 
which iteratively removed $3\sigma$-outliers to improve the fit.
The solution converged to:
\begin{align} 
r_{\rm IPHAS} = r_{\rm APASS} - 0.121 + 0.032(r-i)_{\rm APASS} \label{eqn:apass_r} \\
i_{\rm IPHAS} = i_{\rm APASS} - 0.364 + 0.006(r-i)_{\rm APASS} \label{eqn:apass_i}
\end{align}
The root mean square (rms) residuals of these transformations 
are 0.041 and 0.051, respectively.
The small colour terms in the equations
indicate that the IPHAS and APASS broadband filters 
are very similar.
The transformations include a large fixed offset,
but this is simply due to the fact that 
APASS magnitudes are given in the AB system
and IPHAS uses Vega magnitudes.
Separate transformations were derived for sight-lines 
with varying extinction properties to investigate the robustness
of the transformations with respect to different reddening regimes.
The variations at these different
sight-lines were found to be insignificant.
This is not surprising because heavily-reddened objects 
are naturally scarce at $r<16$.

Having transformed APASS magnitudes into the IPHAS system,
we then computed the median magnitude offset 
for each field which contained at least 30 cross-matched stars.
This was achieved for 48\% of our fields.
The mean offset was found to be
$0.014\pm0.104$~mag in $r$ and $0.007\pm0.108$~mag in $i$
(Table~\ref{tbl:offsets_before}).
A total of 4596 fields showed a median offset
exceeding $\pm$0.03~mag in either $r$ or $i$.

@@

\begin{figure*}
    \includegraphics[width=\textwidth]{plots/calibration/APASS-IPHAS-DR2_rshift.pdf} 
    \includegraphics[width=\textwidth]{plots/calibration/colourbar_apass_r.pdf} 
    \caption{Median magnitude offset in the $r$ band between IPHAS and APASS,
             plotted on a field-by-field basis
             prior to the re-calibration procedure.
             Each square represents the footprint of an IPHAS field
             which contains at least 30 stars with a counterpart
             in the APASS DR7 catalogue.
             The colours denote the median
             IPHAS-APASS magnitude offset in each field,
             which was computed after applying the APASS-to-IPHAS
             transformation to the APASS magnitudes (Eqn.~\ref{eqn:apass_r}).
             For clarity, we do not show the fields at the offset positions.}
        \label{fig:apass_r}
    \vspace{1cm}
    \includegraphics[width=\textwidth]{plots/calibration/SDSS-IPHAS_rshift.pdf}
    \includegraphics[width=\textwidth]{plots/calibration/colourbar_sdss_r.pdf} 
    \caption{Median magnitude offset in the $r$ band
             between IPHAS and SDSS after the re-calibration
             procedure was applied.
             Each square represents the footprint of an IPHAS field
             which contains at least 30 stars
             with a counterpart in the SDSS DR9 catalogue.
             The colours denote the median IPHAS-SDSS magnitude offset
             in each field,
             which was computed after applying the SDSS-to-IPHAS
             transformation to the SDSS magnitudes (Eqn.~\ref{eqn:sdss_r}).
             For clarity, we do not show the fields at the offset positions.}
    \label{fig:sdss_r}
\end{figure*}

\begin{figure*}
	\vspace{2cm}
    \includegraphics[width=\textwidth]{plots/calibration/APASS-IPHAS-DR2_ishift.pdf} 
    \includegraphics[width=\textwidth]{plots/calibration/colourbar_apass_i.pdf} 
    \caption{Same as Fig.~\ref{fig:apass_r} for the $i$-band.}
        \label{fig:apass_i}
    \vspace{1cm}
    \includegraphics[width=\textwidth]{plots/calibration/SDSS-IPHAS_ishift.pdf}
    \includegraphics[width=\textwidth]{plots/calibration/colourbar_sdss_i.pdf} 
    \caption{Same as Fig.~\ref{fig:sdss_i} for the $i$-band.}
    \label{fig:sdss_i}
\end{figure*}

\begin{table}
    \caption{Mean magnitude offsets for objects
             cross-matched between IPHAS and APASS/SDSS
             before the uniform re-calibration.
             Eqns.~\ref{eqn:apass_r}-\ref{eqn:sdss_i} were applied
             to the APASS/SDSS magnitudes to bring them into the
             Vega-based IPHAS system prior to computing
             the offsets.)
             }
    \label{tbl:offsets_before}
    \begin{center}
        \begin{tabular}{lcc}
            \toprule
            {\bf Before re-calibration} & Mean & $\sigma$  \\
            \midrule
            $r$ (IPHAS - APASS) & +0.014 & 0.104 \\
            $i$ (IPHAS - APASS) & +0.007 & 0.108 \\
            $r$ (IPHAS - SDSS) & +0.016 & 0.088 \\
            $i$ (IPHAS - SDSS) & +0.010 & 0.089 \\
            \bottomrule
       \end{tabular}
       \caption{Same as Table~\ref{tbl:offsets_before}
                but computed after the uniform re-calibration
                 was carried out.}
        \label{tbl:offsets_after}
            \begin{tabular}{lcc}
                \toprule
                {\bf After re-calibration} & Mean & $\sigma$ \\
                \midrule
                $r$ (IPHAS - APASS) & +0.000 & 0.011\\
                $i$ (IPHAS - APASS) & +0.000 & 0.011 \\
                $r$ (IPHAS - SDSS)  & -0.001 & 0.029\\
                $i$ (IPHAS - SDSS) & -0.002 & 0.032 \\
                \bottomrule
            \end{tabular}
        \end{center}
\end{table}

@@

At the time of writing, the APASS catalogue did not provide 
sufficient coverage for 7359 of the fields in our data release.
Fortunately, these fields are located mainly 
in the early part of the Galactic Plane (Fig.~\ref{fig:apass_r}),
which were typically observed during the summer months
when photometric conditions are more prevalent at the telescope.
These remaining fields have nevertheless 
been brought onto the same uniform scale 
by employing an ubercalibration-style scheme,
which minimises the magnitude offsets between stars
located in the overlap regions between neighbouring fields.
%Although we explained earlier that these overlap regions
%are prone to systematics, the use of APASS enabled us to
%keep the zeropoints of roughly half of the fields fixed,
%which avoids these systematics from combining to introduce
%artificial gradients across the survey.
%In the next section we will show this to be true by validating
%our calibration against SDSS.
An algorithm for achieving this minimisation
%of minimising the magnitude differences between overlapping frames 
has previously been described by~\citet{Glazebrook1994}.
In brief, there are two fundamental quantities to be
minimised between each pair of overlapping exposures, 
denoted by the indices $i$ and $j$. 
Firstly, the mean magnitude difference between stars in the overlap
region $\Delta_{ij}=\langle m_i-m_j\rangle=-\Delta_{ji}$ is a local
constraint. 
Secondly, to ensure the solution does not stray far 
from the existing calibration, 
the difference in zeropoints 
$\Delta\mathrm{ZP}_{ij}=-\Delta\mathrm{ZP}_{ji}$ 
between each pair of exposures must also be minimised.

Minimisation of these two quantities is a linear least squares problem 
because the magnitude $m$ depends linearly on the ZP (Eqn.~\ref{eqn:mag}).
Hence we can find the ZP shift to be applied to each field 
by minimising the sum:
\begin{equation}
   S = \sum_{i=1}^N \sum_{j=1}^N w_{ij} \theta_{ij} (\Delta_{ij} + a_i - a_j)^2
   \label{eqn:chi2}
\end{equation}
where $i$ denotes the exposure of interest, 
$j$ an overlapping exposure, 
$N$ the number of exposures,
$a_i$ the ZP to solve for 
and $a_j$ the ZP of an overlapping field ($\Delta\mathrm{ZP}_{ij}=a_i-a_j$). 
$w_{ij}$ are weights set to the inverse square of the uncertainty in $\Delta_{ij}$
and $\theta_{ij}$ is an overlap function 
equal to either 1 if exposures $i$ and $j$ overlap or 0 otherwise. 
Solving for $a_i$ is equivalent to solving $\partial
S/\partial a_i=0$, which gives the matrix equation:
\begin{equation}
   \sum_{j=1}^N A_{ij} a_j = b_j,
   \label{eqn:matrix}
\end{equation}
where 
\begin{eqnarray}
   A_{ij} &=& \delta_{ij} \sum_{k=1}^N w_{jk}\theta_{jk} - w_{ij} \theta_{ij},\label{eqn:aij}\\
   b_i &=& \sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ji} = -\sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ij}.\label{eqn:bi}
\end{eqnarray}
This prescription is essentially identical to \citet{Glazebrook1994}.
%,although we changed the formulation to ensure that $A_{ij}$ is positive definite.

@@

Having transformed SDSS magnitudes into the IPHAS system,
we then computed the median magnitude offset for each IPHAS field
which contained at least 30 objects with a cross-matched counterpart
in SDSS.  This was the case for 2\,602 fields.
The median offsets for each of these fields
are shown in Figs.~\ref{fig:sdss_r}-\ref{fig:sdss_i}.
Importantly, the mean offset and standard deviation found 
is $-0.001\pm0.029$~mag in $r$
and $-0.002\pm0.032$~mag in $i$ (Table~\ref{tbl:offsets_after}).
In comparison, offsets computed in the identical way
\emph{before} carrying out our re-calibration showed means
of $+0.016\pm0.088$~mag in $r$ and $+0.010\pm0.089$~mag in $i$ (Table~\ref{tbl:offsets_after}).
We conclude that our re-calibration procedure has
been successful in improving the
uniformity of the calibration by a factor three
and has achieved our aim of bringing the
accuracy down to the level of 0.03~mag (rms).

We warn that the SDSS comparison revealed a number of fields where the offsets
exceeded 0.05~mag (523 fields) or even 0.1~mag (18 fields).
This pattern of outliers is consistent with the tails of a Gaussian
with mean $\approx0$ and $\sigma=0.03$.  Furthermore, it should not
be forgotten that both the SDSS and APASS calibrations are approximations to
perfection and will not be entirely free of anomalies.  Indeed as we worked, 
we noticed the occasional unsurprising examples of inconsistency between these 
two surveys.  
%, we hope to draw upon
%the PanSTARRS survey \citep{Schlafly2012}
%to further improve the accuracy of our calibration.
%At the time of preparing this work data from PanSTARRS
%had not been made public yet.

@@

\subsection{Photometric uncertainties and repeatability}

Fig.~\ref{fig:uncertainties} shows the mean photometric
uncertainty as a function of magnitude in each band.
We find that the uncertainty typically
reaches 0.1~mag near 20.5 in $r$ 
and 19.5 in $i$/\ha\ for the default 2.3\arcsec\ aperture.
At this point we note that the average colour
across all sources in the catalogue is
$1.06\pm0.12$ for $(r-i)$ and $0.44\pm0.03$ for $(r-\ha)$. The
fainter limiting magnitude in $r$ is hence compensated
by the fact that most sources have 
brighter magnitudes in both $i$ and \ha.

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/uncertainties/uncertainties.pdf} 
    \caption{Mean photometric uncertainties
             for $r$ (top), $i$ (middle) and \ha\ (bottom).
             Data points shown are the average values of
             columns \emph{rErr}, \emph{iErr} and \emph{haErr}
             in the catalogue, 
             and the error-bars show the standard deviations.
             The dashed and solid lines indicate 
             the 10$\sigma$ and 5$\sigma$ limits, respectively.
             These uncertainties are based only on the (Poissonian)
             photon noise and hence this figure does not show
             the level of systematic uncertainties.}
    \label{fig:uncertainties}
\end{figure}

The uncertainties shown in Fig.~\ref{fig:uncertainties}
are the random errors based on the expected Poissonian photon noise.
Systematics, such as calibration and deblending errors,
are not included.
To appraise the extent to which our photometry is affected
by such systematics, we can exploit the
secondary measurements which are present in the catalogue.
Fig.~\ref{fig:pairmag} shows the mean residuals between
the primary and secondary magnitudes
-- i.e. the average difference between catalogue columns \emph{r}-\emph{r2},
\emph{i}-\emph{i2}, \emph{ha}-\emph{ha2} -- as a function of magnitude.
We find that sources across the magnitude range 
$13 < r < 17$ are consistent at the level of 5\%
(i.e. $\sigma_{r-r2} \le 0.05$ mag),
with the best repeatability
reached at $r=14$ ($\sigma_{r-r2}~=~0.041$~mag).
The brightest stars tend to show large residuals
-- e.g. $\sigma_{r-r2}~=~0.14$~mag at $r=12$ --
which is due to saturation effects.
At the faint end we find residuals which
show significantly more scatter than would be expected
from photon noise alone;
the effects of source blending and background subtraction
appear to dominate from $\sim$20th magnitude onward.

\begin{figure*}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/repeatability/repeatability.pdf}
\caption{}
\label{fig:pairmag}
\end{subfigure}
\hspace{0.5cm}
\begin{subfigure}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/repeatability/repeatability-reliable.pdf}
\caption{}
\label{fig:pairmag_reliable}
\end{subfigure}
\caption{Photometric repeatability as a function of magnitude
         for both all sources in the catalogue (left panel)
         and for the \emph{veryReliable} sources alone (right panel).
         The values shown are the mean absolute residuals
         between the primary and the secondary detections,
         while the error-bars show the standard deviations.
         Left panel: 
         the best photometric repeatability is reached at $r=14$
         with $\sigma_{r-r2}~=~0.041$~mag.
         Bright stars at $r<13$ and $i<12$ 
         show increasing uncertainties due to saturation effects.
         Right panel: we find that applying the quality criteria
         has improved the photometric repeatability significantly.
         The best repeatability is again reached at $r=14$
         but has reduced to $\sigma=0.028$ mag.
         The quality criteria have also been successful
         at removing objects with large systematics at the bright
         and faint ends.
    }
\end{figure*}

In Fig.~\ref{fig:pairmag_reliable} we show 
a similar comparison of the primary and secondary detections,
but this time we have only included sources which are flagged
as \emph{veryReliable} in the catalogue
(i.e. not saturated, not confused, not near bright stars, etc.)
We find that the average residuals are significantly better
for this subset. Sources across the magnitude range 
$13 < r < 17$ now show a scatter of 0.03 instead of 0.05~mag.
The best repeatability is again reached at $r=14$,
but this time with $\sigma_{r-r2}~=~0.028$~mag.
We conclude that the \emph{veryReliable} quality criteria are effective
in reducing the systematic errors to
the same level as the estimated accuracy of the
global photometric calibration (cf. \S\ref{sec:calibration}).
Moreover, the large systematics at the bright and faint end
disappear.

@@

The number of sources in our catalogue,
counted in 1-by-10~deg$^2$ constant-longitude strips,
is shown in Fig.~\ref{fig:sourcecount}
as a function of Galactic longitude (thick blue line).	
Unsurprisingly, we find the number of sources
to increase towards the Galactic centre.
For example, the average source density near $l\simeq 30^\circ$
is roughly 300\,000 objects per square degree,
which is six times more than the density
found near $l\simeq 180^\circ$.

In addition to the global trend,
real variations are also apparent on smaller scales.
For example, we find a significant drop near the constellations 
of Aquila ($l\simeq40$) and Cygnus ($l\simeq80$ and $l\simeq90$),
which are regions known to be affected
by high levels of foreground extinction
\citep[the extremities of `the Great Rift', e.g.][]{BokBok}.
However, we warn that the source counts shown 
have not been corrected for field pairs that have yet
to be released or for variations in the depth across the included fields.
For example, the dip in the density near $l\simeq140^\circ$
is an artificial feature caused by gaps
in the footprint coverage (seen in Fig.~\ref{fig:footprint}).

>>>>

Although the IPHAS data do contain a significant number of 
such inter-field repeat measurements,
we have found the information contained
in these regions to be insufficient
to constrain the calibration parameters.
This is because photometry at the extreme edges of the WFC
-- where neighbouring fields overlap -- 
is prone to systematics at the level of 1--2\%.
The cause of these errors is thought to include 
the use of twilight sky flats in the pipeline,
which are known to be imperfect for calibrating stellar photometry 
due to stray light and vignetting \citep[e.g.][]{Manfroid1995}.
Moreover, the illumination correction in the overlap regions
is more affected by a radial geometric distortion in the WFC,
which causes the pixel scale to increase as the edges 
are approached \citep{Gonzalez-Solares2011}.
Although these systematics are reasonably small within a single field,
they can combine, during the global calibration routine, to cause artificial zeropoint gradients 
across the survey unless controlled by other external constraints.

For these reasons, we have decided not to depend
on an ubercalibration-type scheme alone,
and have opted to involve an external reference survey
-- where available --
to bring the majority of our data onto a uniform calibration.

@@

We have been able to benefit from the
AAVSO Photometric All-Sky Survey
(APASS; http://www.aavso.org/apass)
to bring most of the survey 
onto a uniform scale.
Since 2009,
APASS has been using two 20~cm-astrographs
to survey the entire sky down to $\sim$17th magnitude
in five filters which include Sloan $r$ and $i$ \citep{Henden2012}.
The most recent catalogue available 
at the time of preparing this work was APASS DR7,
which provides a good coverage across $\sim$half
of the IPHAS footprint.
The overlap regions are shown in Fig.~\ref{fig:apass_r}.
The photometric accuracy of APASS is currently estimated 
to be at the level of 3\% (Henden, private communication),
which is significantly better 
than the original nightly calibrations of IPHAS
which are accurate to $\sim$10\% (Tab.~\ref{tbl:offsets_before}).
APASS achieves its uniform accuracy 
by measuring each star at least two times in photometric conditions,
along with ample standard fields,
benefiting from the large 3-by-3 degrees field of view of its detectors.

With the aim of bringing IPHAS to a similar accuracy of 3\%,
we used the APASS catalogue to identify and adjust the calibration of all IPHAS fields 
which showed a magnitude offset larger than 3\% against APASS.  Experience of re-running
the calibration and testing the results showed us that it was inadvisable to more
finely tune the match for IPHAS data obtained in what were generally the most photometric
nights.  To this end,
the $r$- and $i$-band detection tables of each IPHAS field
were cross-matched against the APASS DR7 catalogue 
using a maximum matching distance of 1~arcsec.
The magnitude range was limited to
$13<r_{\rm APASS}<16.5$ and $12.5<i_{\rm APASS}<16.0$
in order to avoid sources 
brighter than the IPHAS saturation limit on one hand, 
and to avoid sources near the faint detection limit of APASS 
on the other.

The resulting set of $\sim$220\,000 cross-matched stars were then used 
to derive APASS-to-IPHAS magnitude transformations
using a linear least-squares fitting routine, 
which iteratively removed $3\sigma$-outliers to improve the fit.
The solution converged to:
\begin{align} 
r_{\rm IPHAS} = r_{\rm APASS} - 0.121 + 0.032(r-i)_{\rm APASS} \label{eqn:apass_r} \\
i_{\rm IPHAS} = i_{\rm APASS} - 0.364 + 0.006(r-i)_{\rm APASS} \label{eqn:apass_i}
\end{align}
The root mean square (rms) residuals of these transformations 
are 0.041 and 0.051, respectively.
The small colour terms in the equations
indicate that the IPHAS and APASS broadband filters 
are very similar.
The transformations include a large fixed offset,
but this is simply due to the fact that 
APASS magnitudes are given in the AB system
and IPHAS uses Vega-based magnitudes.
Separate transformations were derived for sight-lines 
with varying extinction properties to investigate the robustness
of the transformations with respect to different reddening regimes.
The variations at these different
sight-lines were found to be insignificant.
This is not surprising because heavily-reddened objects 
are naturally scarce at $r<16$.

Having transformed APASS magnitudes into the IPHAS system,
we then computed the median magnitude offset 
for each field which contained at least 30 cross-matched stars.
This was achieved for 48\% of our fields (shown in Figs.~\ref{fig:apass_r} and \ref{fig:apass_i}).
The mean offset was found to be
$0.014\pm0.104$~mag in $r$ and $0.007\pm0.108$~mag in $i$
(Table~\ref{tbl:offsets_before}).
A total of 4596 fields showed a median offset
exceeding $\pm$0.03~mag in either $r$ or $i$.

@@

\begin{figure*}
    \includegraphics[width=\textwidth]{plots/calibration/APASS-IPHAS-DR2_rshift.pdf} 
    \includegraphics[width=\textwidth]{plots/calibration/colourbar_apass_r.pdf} 
    \caption{Median magnitude offset in the $r$ band between IPHAS and APASS,
             plotted on a field-by-field basis
             prior to the re-calibration procedure.
             Each square represents the footprint of an IPHAS field
             which contains at least 30 stars with a counterpart
             in the APASS DR7 catalogue.
             The colours denote the median
             IPHAS-APASS magnitude offset in each field,
             which was computed after applying the APASS-to-IPHAS
             transformation to the APASS magnitudes (Eqn.~\ref{eqn:apass_r}).
             For clarity, we do not show the fields at the offset positions.}
        \label{fig:apass_r}
    \vspace{1cm}
    \includegraphics[width=\textwidth]{plots/calibration/APASS-IPHAS-DR2_ishift.pdf} 
    \includegraphics[width=\textwidth]{plots/calibration/colourbar_apass_i.pdf} 
    \caption{Same as Fig.~\ref{fig:apass_r} for the $i$-band.}
    \label{fig:apass_i}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{plots/calibration/SDSS-IPHAS_rshift.pdf}
    \includegraphics[width=\textwidth]{plots/calibration/colourbar_sdss_r.pdf} 
    \caption{Median magnitude offset between IPHAS and SDSS in the $r$ band
             \emph{after} the re-calibration procedure was applied.
             Each square represents the footprint of an IPHAS field
             which contains at least 30 stars
             with a counterpart in the SDSS DR9 catalogue.
             The colours denote the median IPHAS-SDSS magnitude offset
             in each field,
             which was computed after applying the SDSS-to-IPHAS
             transformation to the SDSS magnitudes (Eqn.~\ref{eqn:sdss_r}).}
    \label{fig:sdss_r}
    \vspace{1cm}
    \includegraphics[width=\textwidth]{plots/calibration/SDSS-IPHAS_ishift.pdf}
    \includegraphics[width=\textwidth]{plots/calibration/colourbar_sdss_i.pdf} 
    \caption{Same as Fig.~\ref{fig:sdss_r} for the $i$-band.}
    \label{fig:sdss_i}
\end{figure*}

\begin{table}
    \caption{Mean magnitude offsets for objects
             cross-matched between IPHAS and APASS/SDSS
             \emph{before} the uniform re-calibration was carried out.
             Transformations were applied
             to the APASS and SDSS magnitudes to bring them into the
             Vega-based IPHAS system prior to computing
             these statistics (see text).
             }
    \label{tbl:offsets_before}
    \begin{center}
        \begin{tabular}{lcc}
            \toprule
            {\bf Before re-calibration} & Mean & $\sigma$  \\
            \midrule
            $r$ (IPHAS - APASS) & +0.014 & 0.104 \\
            $i$ (IPHAS - APASS) & +0.007 & 0.108 \\
            $r$ (IPHAS - SDSS) & +0.016 & 0.088 \\
            $i$ (IPHAS - SDSS) & +0.010 & 0.089 \\
            \bottomrule
       \end{tabular}
       \caption{Same as Table~\ref{tbl:offsets_before}
                but computed \emph{after} the uniform re-calibration
                 was carried out.}
        \label{tbl:offsets_after}
            \begin{tabular}{lcc}
                \toprule
                {\bf After re-calibration} & Mean & $\sigma$ \\
                \midrule
                $r$ (IPHAS - APASS) & +0.000 & 0.011\\
                $i$ (IPHAS - APASS) & +0.000 & 0.011 \\
                $r$ (IPHAS - SDSS)  & -0.001 & 0.029\\
                $i$ (IPHAS - SDSS) & -0.002 & 0.032 \\
                \bottomrule
            \end{tabular}
        \end{center}
\end{table}

@@

At the time of writing, the APASS catalogue did not provide 
sufficient coverage for 7359 of our fields.
Fortunately, these fields are located mainly 
at low Galactic longitudes (cf. Figs.~\ref{fig:apass_r} and \ref{fig:apass_i}),
which were typically observed during the summer months
when photometric conditions are more prevalent at the telescope.
These remaining fields have nevertheless 
been brought onto the same uniform scale 
by employing an ubercalibration-style scheme,
which minimises the magnitude offsets between stars
located in the overlap regions between neighbouring fields.

An algorithm for achieving this minimisation
has previously been described by~\citet{Glazebrook1994}.
In brief, there are two fundamental quantities to be
minimised between each pair of overlapping exposures, 
denoted by the indices $i$ and $j$. 
Firstly, the mean magnitude difference between stars in the overlap
region $\Delta_{ij}=\langle m_i-m_j\rangle=-\Delta_{ji}$ is a local
constraint. 
Secondly, to ensure the solution does not stray far 
from the existing calibration, 
the difference in zeropoints 
$\Delta\mathrm{ZP}_{ij}=-\Delta\mathrm{ZP}_{ji}$ 
between each pair of exposures must also be minimised.

Minimisation of these two quantities is a linear least squares problem 
because the magnitude $m$ depends linearly on the ZP (Eqn.~\ref{eqn:mag}).
Hence we can find the ZP shift to be applied to each field 
by minimising the sum:
\begin{equation}
   S = \sum_{i=1}^N \sum_{j=1}^N w_{ij} \theta_{ij} (\Delta_{ij} + a_i - a_j)^2,
   \label{eqn:chi2}
\end{equation}
where $i$ denotes an exposure, 
$j$ an overlapping exposure, 
$N$ the number of exposures,
$a_i$ the ZP to solve for 
and $a_j$ the ZP of an overlapping field ($\Delta\mathrm{ZP}_{ij}=a_i-a_j$). 
$w_{ij}$ are weights set to the inverse square of the uncertainty in $\Delta_{ij}$,
and $\theta_{ij}$ is an overlap function 
equal to either 1 if exposures $i$ and $j$ overlap or 0 otherwise. 
Solving for $a_i$ is equivalent to solving $\partial
S/\partial a_i=0$, which gives the matrix equation:
\begin{equation}
   \sum_{j=1}^N A_{ij} a_j = b_j,
   \label{eqn:matrix}
\end{equation}
where 
\begin{eqnarray}
   A_{ij} &=& \delta_{ij} \sum_{k=1}^N w_{jk}\theta_{jk} - w_{ij} \theta_{ij},\label{eqn:aij}\\
   b_i &=& \sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ji} = -\sum_{j=1}^N w_{ij} \theta_{ij}\Delta_{ij}.\label{eqn:bi}
\end{eqnarray}

@@

Having transformed SDSS magnitudes into the IPHAS system,
we then computed the median magnitude offset for each IPHAS field
which contained at least 30 objects with a cross-matched counterpart
in SDSS.  This was the case for 2\,602 fields.
The median offsets for each of these fields
are shown in Figs.~\ref{fig:sdss_r} and \ref{fig:sdss_i}.
Importantly, the mean offset and standard deviation found 
is $-0.001\pm0.029$~mag in $r$
and $-0.002\pm0.032$~mag in $i$ (Table~\ref{tbl:offsets_after}).
In comparison, offsets computed in the identical way
\emph{before} carrying out our re-calibration showed means
of $+0.016\pm0.088$~mag in $r$ and $+0.010\pm0.089$~mag in $i$ (Table~\ref{tbl:offsets_after}).
We conclude that our re-calibration procedure has
been successful in improving the
uniformity of the calibration by a factor three
and has achieved our aim of bringing the
accuracy down to the level of 0.03~mag (rms).

We warn that the SDSS comparison revealed a number of fields where the offsets
exceeded 0.05~mag (523 fields) or even 0.1~mag (18 fields).
This pattern of outliers is consistent with the tails of a Gaussian distribution
with $\sigma=0.03$.  Furthermore, it should not
be forgotten that both the SDSS and APASS calibrations are approximations to
perfection and will not be entirely free of anomalies.  Indeed as we worked, 
we noticed the occasional unsurprising examples of inconsistency between these 
two surveys.

@@

\subsection{Random and systematic uncertainties}

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/uncertainties/uncertainties.pdf} 
    \caption{Mean photometric uncertainties
             for $r$ (top), $i$ (middle) and \ha\ (bottom).
             Data points shown are the average values of
             columns \emph{rErr}, \emph{iErr} and \emph{haErr}
             in the catalogue 
             and the error-bars show the standard deviations.
             The dashed and solid lines indicate 
             the 10$\sigma$ and 5$\sigma$ limits, respectively.
             These statistics show the average level of the Poissonian
             photon noise and do not include systematic uncertainties.}
    \label{fig:uncertainties}
\end{figure}

Fig.~\ref{fig:uncertainties} shows the mean photometric
uncertainties (\emph{rErr}, \emph{iErr}, \emph{haErr})
as a function of magnitude.
We find that the uncertainty typically
reaches 0.1~mag near 20.5 in $r$ 
and 19.5 in $i$ and \ha.
We note that the fainter depth in $r$ is compensated
by the fact that most stars have brighter magnitudes in $i$ and \ha;
the average colours in the catalogue are
$\overline{(r-i)}=1.06\pm0.12$ and $\overline{(r-\ha)}=0.44\pm0.03$.
We warn that the statistics shown in Fig.~\ref{fig:uncertainties}
are the random errors based on the expected Poissonian photon noise.
Systematics, such as calibration and deblending errors,
are not included.

To appraise the level at which our photometry is affected by systematics,
we can exploit the secondary measurements which are present in the catalogue
(i.e. \emph{r} vs \emph{r2}, \emph{i} vs \emph{i2}, \emph{ha} vs \emph{ha2}).
In Fig.~\ref{fig:pairmag} we show the mean absolute residuals between
these primary and secondary magnitudes as a function of magnitude (black dots).
We also plot the Poissonian uncertainties for comparison (solid red line).
We find the mean residual to equal 0.03~mag
across the magnitude ranges 13 to 18 ($r$)
and 12 to 17 ($i$, \ha),
consistent with the accuracy of our calibration estimated earlier.
Stars fainter than this range appear to be dominated by photon noise (red line),
while stars at the bright end appear to suffer from significant systematic
errors due to saturation effects.

\begin{figure*}
\captionsetup[subfigure]{labelformat=empty}
\begin{subfigure}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/repeatability/repeatability.pdf}
\caption{}
\label{fig:pairmag}
\end{subfigure}
\hspace{0.5cm}
\begin{subfigure}[b]{0.45\linewidth}
\centering
\includegraphics[width=\textwidth]{figures/repeatability/repeatability-reliable.pdf}
\caption{}
\label{fig:pairmag_reliable}
\end{subfigure}
\caption{Photometric repeatability as a function of magnitude
         for all sources in the catalogue (left panel)
         and for the \emph{veryReliable} sources alone (right panel).
         Black dots show the mean absolute residuals
         between the primary and the secondary detections.
         The error-bars show the standard deviations.
         The red trend line shows the average Poissonian uncertainties
         we derived in Fig.~\ref{fig:uncertainties}.
         We find that the \emph{veryReliable} quality criteria are successful
         at removing objects with large uncertainties.
    }
\end{figure*}

In Fig.~\ref{fig:pairmag_reliable} we show 
a similar comparison between the primary and secondary detections,
but this time we have only included sources which are flagged
as \emph{veryReliable} in the catalogue.
We do not observe an improvement in the average residuals as a function of magnitude,
but the number of outliers is significantly smaller
(evidenced by the shorter error bars which denote the standard deviation of the absolute residuals).
We conclude that the \emph{veryReliable} quality criteria are effective
at reducing the level of the systematic errors,
while also removing the unreliable data at the bright and faint end.

@@

Fig.~\ref{fig:sourcecount} shows the number of sources
in the catalogue counted in $1^\circ$-wide strips
as a function of Galactic longitude (thick blue line).	
Unsurprisingly, we find the number of sources
to increase towards the Galactic centre.
For example, the average source density near $l\simeq 30^\circ$
is roughly 300\,000 objects per square degree,
which is six times more than the density
found near $l\simeq 180^\circ$.
In addition to the global trend,
variations are also apparent on smaller scales.
For example, we find a significant drop near the constellations 
of Aquila ($l\simeq40$) and Cygnus ($l\simeq80$ and $l\simeq90$),
which are regions known to be affected
by high levels of foreground extinction
\citep[the extremities of `the Great Rift', e.g.][]{BokBok}.
However, we warn that the source counts shown 
have not been corrected for field pairs that have yet
to be released or for variations in the depth across the included fields.
For example, the dip in the density near $l\simeq140^\circ$
is an artificial feature caused by gaps
in the footprint coverage (seen in Fig.~\ref{fig:footprint}).
